---
output: github_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
library(mrangr)
library(terra)

load("inst/extdata/readme/first_com.rda")
load("inst/extdata/readme/first_sim.rda")

```

# mrangr

<!-- badges: start -->

[![CRAN release](https://www.r-pkg.org/badges/version/mrangr?color=green)](https://cran.r-project.org/package=mrangr) [![Project Status: Active -- The project has reached a stable, usable state and is being actively developed.](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)

<!-- badges: end -->

The **mrangr** package is designed to simulate metacommunities within a **spatially explicit, mechanistic framework**. It extends the functionality of the [`rangr`](https://github.com/ropensci/rangr) package by allowing for the simulation of **multiple interacting species** via an asymmetric interaction matrix.

This tool mimics the essential processes shaping metacommunity dynamics: local population growth, dispersal, and interspecific interactions. Simulations take place in dynamic environments, facilitating projections of community shifts in response to environmental changes.

## Installation

You can install **mrangr** with:

```{r intall, eval=FALSE}
install.packages("mrangr")
```

## Basic Workflow

The `mrangr` workflow involves initialising a community with spatial data and interaction parameters, running the simulation, and analysing the results.

### 1. Input Maps and Interactions

You must provide carrying capacity maps (`K_map`) and initial abundance maps (`n1_map`) as `SpatRaster` objects. For a community of $N$ species, the rasters must contain $N$ layers.

```{r kmap}
# Load example maps
K_map <- rast(system.file("input_maps/K_map_eg.tif", package = "mrangr"))
K_map <- subset(K_map, 1:2)
```


Interspecific interactions are defined using an **interaction matrix** ($a$), where values represent the per-capita interaction strength of the species in the column on the species in the row.

```{r inter}
# Example for 2 species with symmetric competition
nspec <- 2
a <- matrix(c(NA, -0.8, -0.8, NA), nrow = nspec, ncol = nspec)
```

### 2. Community Initialisation

Use `initialise_com()` to create a `sim_com_data` object. This stores all parameters, including the intrinsic growth rate ($r$) and the dispersal rate.

```{r init, eval=FALSE}
first_com <- initialise_com(
  n1_map = round(K_map / 2), 
  K_map = K_map, 
  r = 1.1, 
  a = a, 
  rate = 1 / 500
)
```

### 3. Running the Simulation

The `sim_com()` function executes the simulation over a specified number of time steps.

```{r sim, eval=FALSE}
first_sim <- sim_com(first_com, time = 100)
```

### 4. Visualisation

You can visualise the final spatial distributions or the change in mean abundance over time.

```{r vis}
# Visualise spatial niches at specific time steps
plot(first_sim, time = c(1, 10, 100))
# Plot abundance time series for all species
plot_series(first_sim)
```

## Virtual Ecologist

The package includes a `virtual_ecologist()` function to simulate real-world observation processes. This allows users to sample the simulated community at defined points in space and time, incorporating sampling effort and detection probability into the simulation.

## Citation

To cite **mrangr**, please use the `citation()` function:

```{r cite, eval=FALSE}
library(mrangr)
citation("mrangr")
```
